### 2014-07-21 Algorithms/Givens/Givens.tex
minor edits to givens lab

Because Givens rotations only operate on two specific rows at a time, they provide multiple ways to iterate over the array we are processing.
For example, using Givens rotations you could, starting at the bottom of each column and continuing up until you reach the main diagonal, zero out each entry by applying a Givens rotation to each row and the row immediately above it.
Alternatively, within each column, you could could zero out the first entry of each row below the main diagonal by rotating it with the row corresponding to the main diagonal.
Yet, another way to do it would be to start at the bottom left corner of the matrix and then zero out each entry from left to right along each diagonal.
This flexibility is what makes the use of Givens rotations ideal for some problems.
When working with sparse matrices, Givens rotations allow for the solution of linear systems through a series of orthonormal operations that operate only on small parts of the array.
The fact that Givens rotations operate only on small portions of the array also lends well to parallel solutions to systems of equations (or least squares problems).
They allow for the sort of flexibility we had when using Gaussian elimination while still maintaining the favorable stability that comes with using orthonormal transformations.

@@

\begin{problem}
\label{prob:givens_hessenberg}
Write a modified version of your solution to Problem \ref{prob:Givens}, call it \li{givens2} that computes the 
QR decomposition of an upper Hessenberg matrix.
Instead of starting at the bottom of each column and working up, just run down the first subdiagonal from left to right.
When operating on $Q$, you do not need to operate on the full length of each row.
It is sufficient to perform the matrix multiplication on the portion of the array \li{Q[j:j+2,:j+2]} where \li{j} is the current column.

@@

Notice how $Q$ in the QR decomposition of an upper Hessenberg matrix is also upper Hessenberg.
What is the computational order of  complexity for this problem?
Approximately for what $m$ is your implementation as fast as the general QR decomposition built in to \li{scipy.linalg} for computing the QR decomposition of an upper Hessenberg matrix?
\end{problem}

\begin{problem}
\label{prob:givens_hessenberg_modified}
You may have noticed that matrix multiplication by $Q$ is generally a $\mathcal{O} \left( n^3 \right)$ algorithm, while the application of these individual Givens rotations is a $\mathcal{O} \left( n^2 \right)$ algorithm.
Write a modified version of your solution to Problem \ref{prob:givens_hessenberg} called \li{givens2_mod} which returns 
an $(n-1) \times 2 \times 2$ array containing the computed values for $G$ at each step in the algorithm in the order in which they are applied to the upper Hessenberg array $H$.

Write two more functions \li{apply_Q} and \li{apply_QT} which, using the matrix of Givens rotations, perform left multiplication 
by $Q$ and $Q^{-1}$, respectively, on some other input array $B$.
Left multiplication by $Q^{-1}$ can be done by applying each of the Givens rotations to $B$ the same way you did to $H$ to compute its QR factorization.
Left multiplication by $Q$ can be done by applying the transpose of the Givens rotations to their corresponding portions of $B$, but in the reverse order.
Notice that you will have to apply each Givens rotation across the full width of the rows it operates on since you do not know anything about the content of $B$.

@@

An interesting side-note is that each Givens rotation can be represented as a single floating point number, so, when operating in place, $Q$ can be stored entirely in the lower triangular portion of the array on which we are operating by storing each rotation in the entry that it zeroes out.
A similar approach would to store Householder reflectors in the columns they zero out.
In either case, we can represent the QR decomposition of an array using only the memory that was originally used to store the array itself.
This is similar to the approach  for computing the LU decomposition entirely in place.
These representations of $Q$ and $R$ can be used in various ways to perform matrix multiplication by $Q$, $Q^T$ and $R$ as needed.

@@

We are now changing the subject to talk about line fitting.

@@

\section*{Loading Data from .npz Files}
For Least Squares problems as well as in many other contexts, loading data is often a necessary step before
proceeding with further analysis. Here we briefly review another data format in Python and the commands used
to load the data.

A \li{.npz} file is a compressed binary file that contains an archive of NumPy data structures.
A given file may therefore contain several arrays, each array associated with a unique string that identifies it.
When you load a \li{.npz} file in Python, a dictionary-like object is returned, and you can access the data by
providing the appropriate key. Note that when you load a \li{.npz} file, you must also be sure to close it when
you are finished. This is taken care of automatically if you use the \li{with ... as} keywords.

@@

We enter them into Python as a $9\times 2$ array:
\begin{lstlisting}
>>> P = np.array([[134,76],[104,146],[34,176],[-36,146],
                  [-66,76],[-36,5],[34,-24],[104,5],[134,76]])
\end{lstlisting}
We compute $A$ and $b$ by entering the following:
\begin{lstlisting}
>>> A = A =np.hstack((2*P, np.ones((9,1))))
>>> b = (P**2).sum(axis=1)
\end{lstlisting}
Hence, we get the least squares solution
\begin{lstlisting}
>>> x = np.dot(np.dot(la.inv(np.dot(A.T,A)),A.T),b)
\end{lstlisting}
Then we find $c_1$, $c_2$, and $r$ by:
\begin{lstlisting}
>>> from math import sqrt
>>> c1, c2, c3 = x
>>> r = sqrt(c1**2 + c2**2 + c3)
\end{lstlisting}
We plot this by executing
\begin{lstlisting}
>>> theta = np.linspace(0,2*np.pi,200)
>>> plt.plot(r*np.cos(theta)+c1,r*np.sin(theta)+c2,'-',P[:,0],P[:,1],'*')
>>> plt.show()
\end{lstlisting}

>>>>

Because Givens rotations only operate on two specific rows at a time, they provide multiple ways to iterate over the array we are processing. Here are three examples showcasing different iterating processes:

\begin{enumerate}
\item Starting at the bottom of each column and continuing up until you reach the main diagonal, zero out each entry by applying a Givens rotation to each row and the row immediately above it.

\item Starting within each column, zero out the first entry of each row below the main diagonal by rotating it with the row corresponding to the main diagonal.

\item Starting at the bottom left corner of the matrix, zero out each entry from left to right along each diagonal.

\end{enumerate}

This flexibility is what makes using Givens rotations ideal for some problems. When working with sparse matrices, Givens rotations allow for the solution of linear systems through a series of orthonormal operations that operate only on small portions of the array. The fact that Givens rotations operate only on small portions of the array also lends well to parallel solutions to systems of equations (or least squares problems).
They allow for the sort of flexibility we had when using Gaussian elimination while still maintaining the favorable stability that comes with using orthonormal transformations.

@@

\begin{problem}
\label{prob:givens_hessenberg}
Write a modified version of your solution to Problem \ref{prob:Givens} called \li{givens2} that computes the QR decomposition of an upper Hessenberg matrix.
Instead of starting at the bottom of each column and working up, just run down the first subdiagonal from left to right.
When operating on $Q$, you do not need to operate on the full length of each row.
It is sufficient to perform the matrix multiplication on the portion of the array \li{Q[j:j+2,:j+2]} where \li{j} is the current column.

@@

Notice how $Q$ in the QR decomposition of an upper Hessenberg matrix is also upper Hessenberg.
What is the computational order of  complexity for this problem?
For approximately what $m$ is your implementation as fast as the general QR decomposition built into \li{scipy.linalg} for computing the QR decomposition of an upper Hessenberg matrix?
\end{problem}

\begin{problem}
\label{prob:givens_hessenberg_modified}
You may have noticed that matrix multiplication by $Q$ is generally a $\mathcal{O} \left( n^3 \right)$ algorithm even though the application of these individual Givens rotations is a $\mathcal{O} \left( n^2 \right)$ algorithm.

Now write a modified version of your solution to Problem \ref{prob:givens_hessenberg} called \li{givens2_mod} which returns 
an $(n-1) \times 2 \times 2$ array containing the computed values for $G$ at each step in the algorithm in the order in which they are applied to the upper Hessenberg array $H$.
\end{problem}

\begin{problem}
Write two more functions \li{apply_Q} and \li{apply_QT} which, using the matrix of Givens rotations, perform left multiplication 
by $Q$ and $Q^{-1}$, respectively, on some other input array $B$.
Left multiplication by $Q^{-1}$ can be done by applying each of the Givens rotations to $B$ the same way you did to $H$ to compute its QR factorization.
Left multiplication by $Q$ can be done by applying the transpose of the Givens rotations to their corresponding portions of $B$, but in the reverse order.
Notice that you will have to apply each Givens rotation across the full width of the rows it operates on since you do not know anything about the content of $B$.

@@

An interesting side-note is that each Givens rotation can be represented as a single floating point number, so, when operating in place, $Q$ can be stored entirely in the lower triangular portion of the array on which we are operating by storing each rotation in the entry that it zeroes out.
A similar approach would be to store Householder reflectors in the columns they zero out. In either case, we can represent the QR decomposition of an array using only the memory that was originally used to store the array itself.
This is similar to the approach  for computing the LU decomposition entirely in place. These representations of $Q$ and $R$ can be used in various ways to perform matrix multiplication by $Q$, $Q^T$, and $R$ as needed.

@@

Let's now talk about line fitting.

@@

\section*{Loading Data from .npz Files}
For Least Squares problems as well as in many other contexts, loading data is often a necessary step before proceeding with further analysis. We will briefly review another data format in Python and the commands used to load the data.

A \li{.npz} file is a compressed binary file that contains an archive of NumPy data structures. A given file may therefore contain several arrays, each array associated with a unique string that identifies it.
When you load a \li{.npz} file in Python, a dictionary-like object is returned and you can access the data by providing the appropriate key. Note that when you load a \li{.npz} file, you must also be sure to close it when
you are finished. This is taken care of automatically if you use the \li{with ... as} keywords.

@@

We enter them into Python as a $9\times 2$ array:
\begin{lstlisting}
>>> P = np.array([[134,76],[104,146],[34,176],[-36,146],
                  [-66,76],[-36,5],[34,-24],[104,5],[134,76]])
\end{lstlisting}
We compute $A$ and $b$ by entering the following:
\begin{lstlisting}
>>> A = np.hstack((2*P, np.ones((9,1))))
>>> b = (P**2).sum(axis=1)
\end{lstlisting}
Hence, we get the least squares solution
\begin{lstlisting}
>>> x = np.dot(np.dot(la.inv(np.dot(A.T,A)),A.T),b)
\end{lstlisting}
Then we find $c_1$, $c_2$, and $r$ by:
\begin{lstlisting}
>>> from math import sqrt
>>> c1, c2, c3 = x
>>> r = sqrt(c1**2 + c2**2 + c3)
\end{lstlisting}
We plot this by executing
\begin{lstlisting}
>>> theta = np.linspace(0,2*np.pi,200)
>>> plt.plot(r*np.cos(theta)+c1,r*np.sin(theta)+c2,'-',P[:,0],P[:,1],'*')
>>> plt.show()
\end{lstlisting}
